---
title: "Project planning: From experiment design to manuscript and data release"
author: "Shravan Vasishth (vasishth.1@osu.edu)"
date: '```r format(Sys.Date(), "%B %d, %Y")```'
output:
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "dove"
    fonttheme: "structurebold"    
    includes:
      in_header: preamble.tex    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Motivation

## Important problems in science today

Results from published papers:

- non-replicable 
- non-reproducible 
- contain serious mistakes
- claims can't be verified
- Impossible to build on existing data (meta-analysis)

## Important problems in science today

- Experiments are run without much of a workflow
- The result is a chaotic research process
- The resulting errors waste researcher's time and energy

# Experiment design, and planning sample size using simulated data

## Fake-data simulation

```{r}

```

# Defining the analysis plan using simulated data 

# Pre-registering the analysis

# Check that your experiment software actually collects the data you need

# Once data are collected, visualize and summarize the data before doing any analysis

# Create an R package to document and release your data and analyses

# Code refactoring

# Integrating the data analysis into the manuscript

# Releasing data and code: a suggested checklist

