---
title: "Title of paper"
shorttitle        : "Short title"
author: 
  - name          : "Your name"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Human Science Faculty, Department of Linguistics, University of Potsdam, 14476 Potsdam, Germany"
    email         : "vasishth@uni-potsdam.de"
  - name          : "Author 2"
    affiliation   : "2"
  - name          : "Author 3"
    affiliation   : "1"
  - name          : "Author 4"
    affiliation   : "1"
  - name          : "Author 5"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Department of Linguistics, University of Potsdam"
  - id            : "2"
    institution   : "Department of Artificial Intelligence, University of Utrecht"

authornote: |

  Funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Project number 317633480 – SFB 1287, project Q.

abstract: |
  abstract here.    
  
keywords          : "aphasia, cue-based retrieval, sentence processing, Bayesian cognitive modeling; k-fold cross validation"
wordcount         : "7000"
bibliography      : ["examplebibliography.bib"]
biblio-style      : "apa"
lang              : "en-EN" 
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no 
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"  ## doc, jou
##appendix          : "appendix.Rmd"
output:
  papaja::apa6_pdf:
    citation_package: biblatex
    latex_engine: xelatex
biblatexoptions: [date=year,labeldate=year,alldates=year, urldate=short, url=false,eprint=false]     
header-includes: |
  \AtEveryBibitem{\clearfield{month}}
  \AtEveryBibitem{\clearfield{day}}
  \AtEveryBibitem{\clearfield{url}}
  \DeclareLanguageMapping{english}{english-apa}
  \usepackage{gb4e}\noautomath
  \usepackage{todonotes}
  \usepackage[utf8]{inputenc}
  \usepackage{fancyvrb}
---

```{r setup, include = FALSE}
library("papaja")
```

# Introduction

Consider the subject and object relative clauses shown in  (\ref{caplan_ex1}):

\begin{exe}
\ex \begin{xlist}
  \ex\label{SR}{\textbf{Subject Relative (SR):} The boy who hugged the girl chased the woman}
  \ex\label{OR}{\textbf{Object Relative (OR):} The brother who the sister followed kissed the woman}
  \end{xlist}
\label{caplan_ex1}
\end{exe}

@grodner hypothesized that subject relatives are easier to process than object relatives. We revisit their research hypothesis using their data.

\todo[inline]{Load data, preprocess}

\todo[inline]{Carry out analysis}


\newpage


<!--COMMENT: # References -->
```{r create_r-references, cache= FALSE}
# to not overwrite the manual  fixes 
## r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup

